---
title: "Hands-on Exercise 00: Working with tidyverse"
author: "Andrea Yeo"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

## Getting started

Loading tidyverse onto r environment by using the code chunk below

```{r}
pacman:: p_load(tidyverse)
```

## Importing Data

Code chunk below uses read_csv() of readr to import REALIS2019.csv into r environment as a tibble data.frame.

```{r}
realis2019 <- read_csv("data/REALIS2019.csv")
popdata_fat <- read_csv("data/PopData2019_fat.csv")
```

## Pivoting Data

```{r}
#| eval: FALSE #Display code but not run it
#pivot_longer() "lengthens" data, increasing the number of rows and decreasing the number of columns

popdata_long <- popdata_fat %>% #pipe: to combine diff functions tgt
  pivot_longer(c(3:21), #column
               names_to = "Age Group", #column headers to rows
               values_to = "Population") #values to rows

```

```{r}
#| eval: FALSE
write_rds(popdata_long, "rds/popdata_long.rds") #rds: compact r file
```

```{r}
#| echo: FALSE #Run the code but not display it
popdata_long <- read_rds("rds/popdata_long.rds")
```

## Selecting Data

```{r}
realis2019_selected <- realis2019 %>%
  select(`Project Name`,
         `Transacted Price ($)`,
         `Type of Sale`,
         `Unit Price ($ psm)`,
         `Property Type`)
realis2019_selected
```

## Filtering Data using dplyr package

```{r}
realis2019_filtered <- realis2019_selected %>%
  filter (`Property Type` == "Condominium" |
            `Property Type` == "Apartment") %>%
  filter (`Type of Sale` == "New Sale") %>%
  filter (`Unit Price ($ psm)` <= 13000)

realis2019_filtered
```

### Putting all together

```{r}
realis2019_end <- realis2019 %>%
  select(`Project Name`,
         `Transacted Price ($)`,
         `Type of Sale`,
         `Unit Price ($ psm)`,
         `Property Type`)   %>%
  filter (`Property Type` == "Condominium" |
            `Property Type` == "Apartment") %>%
  filter (`Type of Sale` == "New Sale") %>%
  filter (`Unit Price ($ psm)` <= 13000)

realis2019_end
```

::: callout-tip
## Reflection

-   `echo`: controls whether the code is displayed
    -   If `echo` = TRUE (default), the code is shown in the rendered document.
    -   If `echo` = FALSE, the code is hidden, but the output or results (if any) are still included.
-   `eval`: controls whether the code is executed
    -   If `eval` = TRUE (default), the code in the chunk is executed, and its results (e.g., output, plots) are included in the document
    -   If `eval` = FALSE, the code is not executed, and no output or results are included, though the code may still be visible if echo = TRUE.
-   `%\>%`: used to combine multiple functions together
:::

#Change environmental variables if needed
