---
title: "Hands-on Exercise 05d"
author: "Andrea Yeo"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

[With the assistance of ChatGPT]{style="font-size: 14px;"}

## 5. Visual Multivariate Analysis with Parallel Coordinates Plot

### 5.1 Overview

Parallel coordinates plots, invented by [Alfred Inselberg](http://www.math.tau.ac.il/~aiisreal/) in the 1970s, are a visualization technique designed for analyzing multivariate numerical data. They help in comparing multiple variables and identifying relationships, commonly used in academic and scientific fields rather than business settings. As noted by [Stephen Few](https://www.perceptualedge.com/articles/b-eye/parallel_coordinates.pdf) (2006), their strength lies in interactive analysis rather than public presentation. A key application includes characterizing clusters in customer segmentation.

By the end of this hands-on exercise, we will be able to:

-   Plot static parallel coordinates plots using `ggparcoord()` from the [**GGally**](https://www.rdocumentation.org/packages/GGally/versions/2.2.1) package.
-   Create interactive parallel coordinates plots with the [**parcoords**](https://www.rdocumentation.org/packages/MASS/versions/7.3-64/topics/parcoord) package.
-   Generate interactive parallel coordinates plots using the [**parallelPlot**](https://www.rdocumentation.org/packages/parallelPlot/versions/0.4.0) package.

### 5.2 Installing and Launching R Packages

For this exercise, the below R packages will be used:

-   GGally: An extension of ggplot2 that simplifies multivariate data visualization, including parallel coordinates plots.

-    [**parcoords**](https://www.rdocumentation.org/packages/MASS/versions/7.3-64/topics/parcoord):A package for creating interactive parallel coordinates plots using htmlwidgets in R.

-    [**parallelPlot**](https://www.rdocumentation.org/packages/parallelPlot/versions/0.4.0):A tool for generating interactive parallel coordinates plots with enhanced visualization capabilities.

-    [**tidyverse**](https://www.rdocumentation.org/packages/tidyverse/versions/2.0.0):A collection of R packages designed for data manipulation, visualization, and analysis following a consistent syntax.

```{r}
pacman::p_load(GGally, parallelPlot, tidyverse)
```

### 5.3 Data Preparation

```{r}
wh <- read_csv("data/WHData-2018.csv")

spec(wh)
```

::: callout-note
-   This section data is the same dataset used in [Hands-on_Ex05c](Hands-on_Ex05c.html)
:::

### 5.4 Plotting Static Parallel Coordinates Plot

This section covers plotting static parallel coordinates plots using [ggparcoord()](https://ggobi.github.io/ggally/reference/ggparcoord.html) from the **GGally** package.

#### 5.4.1 Plotting a simple parallel coordinates

Code below shows a typical syntax used to plot a basic static parallel coordinates plot by using `ggparcoord()`.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12))
```

-   Only two arguments, `data` and `columns`, are used in `ggparcoord()`.
-   The data argument maps the dataset (e.g., wh), while columns selects the variables for the parallel coordinates plot.

#### 5.4.2 Plotting a parallel coordinates with boxplot

To improve the interpretability of the World Happiness parallel coordinates plot, `ggparcoord()` provides customization options like - groupColumn for grouping, - scale for variable scaling, - alphaLines to reduce the intensity of the line colour, - showPoints for displaying data points, - splineFactor for smoothing, - title for parallel coordinates plot a title, - alphaLines for transparency control, and - boxplot for overlaying distributions.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Parallel Coordinates Plot of World Happines Variables")
```

#### 5.4.3 Parallel coordinates with facet

Since `ggparcoord()` extends **ggplot2**, it allows integration with **ggplot2** functions.

The example below uses `facet_wrap()` to create multiple small parallel coordinates plots, each representing a different geographical region, such as East Asia.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region)
```

One of the aesthetic defect of the current design is that some of the variable names overlap on x-axis.

#### 5.4.4 Rotating x-axis text label

To improve readability, the x-axis labels are rotated by 30 degrees using the `theme()` function in ggplot2. This enhances clarity when displaying parallel coordinates plots.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30))
```

Thing to learn from the code chunk above:

-   To rotate x-axis text labels, use `axis.text.x` in the `theme()` function and set `element_text(angle = 30)` to rotate the labels by 30 degrees.

#### 5.4.5 Adjusting the rotated x-axis text label

To prevent label overlap after rotating x-axis text by 30 degrees, adjust text positioning using the `hjust` argument in `element_text()` within the `theme()` function, applied to `axis.text.x`.

```{r}
ggparcoord(data = wh, 
           columns = c(7:12), 
           groupColumn = 2,
           scale = "uniminmax",
           alphaLines = 0.2,
           boxplot = TRUE, 
           title = "Multiple Parallel Coordinates Plots of World Happines Variables by Region") +
  facet_wrap(~ Region) + 
  theme(axis.text.x = element_text(angle = 30, hjust=1))
```

### 5.5 Plotting Interactive Parallel Coordinates Plot: parallelPlot methods

#### 5.5.1 The basic plot

The [parallelPlot](https://cran.r-project.org/web/packages/parallelPlot/) package in R, built on htmlwidgets and d3.js, enables interactive parallel coordinates plots. This section covers its key functions for creating dynamic visualizations.

```{r}
wh <- wh %>%
  select("Happiness score", c(7:12))
parallelPlot(wh,
             width = 320,
             height = 250)
```

Notice that some of the axis labels are too long.

#### 5.5.2 Rotate axis label

`rotateTitle` argument is used to avoid overlapping axis labels.

```{r}
parallelPlot(wh,
             rotateTitle = TRUE)
```

A key interactive feature of parallelPlot allows users to click on a variable, such as Happiness Score, changing the default blue color to varying intensities for better visualization.

#### 5.5.3 Changing the colour scheme

We can change the default blue colour scheme by using continousCS argument as shown in the code below:

```{r}
parallelPlot(wh,
             continuousCS = "YlOrRd",
             rotateTitle = TRUE)
```

#### 5.5.4 Parallel coordinates plot with histogram

`histoVisibility` argument is used to plot histogram along the axis of each variables.

```{r}
histoVisibility <- rep(TRUE, ncol(wh))
parallelPlot(wh,
             rotateTitle = TRUE,
             histoVisibility = histoVisibility)
```

### 5.6 References

-   Kam, T.S(2024). [Visual Statistical Analysis.](https://r4va.netlify.app/chap15)
-   [ggparcoord()](https://ggobi.github.io/ggally/reference/ggparcoord.html) of [GGally](https://ggobi.github.io/ggally/index.html) package
-   [parcoords user guide](https://plotly.github.io/plotly.py-docs/generated/plotly.graph_objects.Parcoords.html)
-   [parallelPlot](https://cran.r-project.org/web/packages/parallelPlot/parallelPlot.pdf)

### 5.7 Takeaway

::: callout-tip
## Key takeaways

-   Parallel Coordinates Plots: Useful for visualizing multivariate numerical data and identifying relationships.
-   Static Plots: Created using ggparcoord() from GGally, allowing customization with grouping, scaling, and faceting.
-   Improving Readability: Rotate x-axis labels (theme()), adjust label positioning (hjust).
-   Interactive Plots: parallelPlot() enables dynamic visualization with adjustable colors, rotated labels, and histograms.
:::

### 5.8 Further exploration

1.  To explore key variables that affect happiness score

-   To determine the optimal number of clusters for segmenting the dataset, I use the Elbow Method, which evaluates the within-cluster sum of squares (WSS) for different values of k (from 1 to 10).

-   The WSS measures the compactness of clusters, with lower values indicating better-defined clusters. By plotting WSS against the number of clusters, I identify the "elbow point"—the value of k where additional clusters no longer significantly improve the clustering quality.

```{r}
head(wh)
```

```{r}
library(ggplot2)

# Run k-means for different values of k (1 to 10)
wss <- sapply(1:10, function(k) {
  kmeans(wh %>% select(1, 2, 4, 5, 6, 7), centers = k, nstart = 10)$tot.withinss
})

# Plot the Elbow Method graph
qplot(1:10, wss, geom = "line") +
  geom_point() +
  labs(title = "Elbow Method for Optimal k", x = "Number of Clusters", y = "WSS") +
  theme_minimal()

```

-   Based on the Elbow Method, I determine that the optimal number of clusters is k = 3. I then apply k-means clustering to the dataset, selecting key numerical variables such as Happiness Score, GDP per Capita, Social Support, Healthy Life Expectancy, Freedom, and Perceptions of Corruption.

-   The k-means algorithm assigns each country to one of the three clusters, grouping similar countries based on these features.

```{r}
optimal_k <- 3  

wh_cluster <- wh %>%
  select(1, 2, 4, 5, 6, 7) %>%
  kmeans(centers = optimal_k, nstart = 10)

wh$Cluster <- as.factor(wh_cluster$cluster)

```

-   To better understand the clustering results, I use a Parallel Coordinates Plot, a multivariate visualization technique that represents each country as a line spanning multiple axes, corresponding to the selected variables.

-   This visualization provides insights into how different groups of countries compare in terms of happiness-related indicators, highlighting key differences between the clusters.

```{r}
ggparcoord(wh, 
           columns = c(1, 2, 4, 5, 6, 7),  
           groupColumn = "Cluster",  
           scale = "uniminmax",  
           alphaLines = 0.3,  
           title = paste("Parallel Coordinates Plot with k =", optimal_k, "Clusters")) +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))  


```

**Key observations:**

Cluster Differentiation:

-   **Cluster 3 (Blue)**: Represents Middle-Income Countries

    -   This cluster falls between the two extremes, with moderate happiness scores and GDP per capita.

    <!-- -->

    -   More variation in freedom to make life choices and generosity, suggesting diverse political and social environments.

    -   Likely includes emerging economies or nations in transition.

    -   Example: Egypt, Cambodia, Laos

-   **Cluster 2 (Green)**: Cluster 2 (Green) Represents Wealthier and Happier Countries

    -   This cluster consistently scores high on GDP per capita, happiness score, life expectancy, and freedom to make life choices.

    -   These countries also have lower perceptions of corruption, indicating a strong governance system.

    -   Likely includes developed nations with strong economies and social support systems.

    -   Example: Denmark, Finland, and Norway

-   **Cluster 1 (Red)**: Represents Lower-Income Countries

    -   Countries in this cluster have lower GDP per capita, happiness scores, and life expectancy.

    <!-- -->

    -   Higher perceptions of corruption suggest governance and institutional challenges.

    -   These may be developing nations with weaker economies.

    -   Example: Tanzania, South Sudan, Central African Republic
