---
title: "Hands-on Exercise 08b"
author: "Andrea Yeo"
date-modified: "last-modified"
execute:
  echo: true
  eval: true
  warning: false
  freeze: true
---

[With the assistance of ChatGPT]{style="font-size: 14px;"}

## 8 Visualising Geospatial Point Data

### 8.1 Overview

Proportional symbol maps (or graduated symbol maps) use symbol size to represent differences in magnitude for discrete data, such as counts of people.

-   Classed maps (range-graded or graduated symbols) group data into categories.

-   Unclassed maps (proportional symbols) scale symbols directly to attribute values.

In this exercise, we will learn to create a proportional symbol map using **tmap** to visualize the number of wins by Singapore Pools' outlets.

#### 8.1.1 Learning outcome
By the end of this exercise, you will learn to:

-   Import an aspatial data file into R.
-   Convert it into a simple point feature data frame and assign a projection reference.
-   Create interactive proportional symbol maps using tmap.

### 8.2 Loading R packages and Data preparation

Before we get started, we need to ensure that tmap package of R and other related R packages have been installed and loaded into R.

```{r}
pacman::p_load(sf, tmap, tidyverse)
```

#### 8.2.1 Data

The dataset *SGPools_svy21.csv* contains seven columns, including XCOORD and YCOORD, which represent the SVY21 projected coordinates of Singapore Pools' outlets and branches.

#### 8.2.2 Data import and preparation

The code below uses `read_csv()` function of **readr** package to import *SGPools_svy21.csv* into R as a tibble data frame called sgpools.

```{r}
sgpools <- read_csv("data/data/aspatial/SGPools_svy21.csv")
```
Next, we examine if the data file has been imported correctly.

```{r}
list(sgpools)
```

#### 8.2.3 Creating a sf data frame from an aspatial data frame

The code below converts *sgpools* data frame into a simple feature data frame by using `st_as_sf()` of **sf** packages

```{r}
sgpools_sf <- st_as_sf(sgpools, 
                       coords = c("XCOORD", "YCOORD"),
                       crs= 3414)
```

-   `coords argument`: Specify the x-coordinate column first, followed by the y-coordinate column.
-   `crs` argument: Defines the coordinate reference system (CRS) using EPSG codes.
  -   [EPSG: 3414](https://epsg.io/3414) is used for the Singapore SVY21 Projected Coordinate System.
-   EPSG codes for other locations can be found at [epsg.io](https://epsg.io/).

We can display the basic information of the newly created sgpools_sf below.

```{r}
list(sgpools_sf)
```

The output confirms that sgppols_sf is a point feature class with EPSG: 3414 (Singapore SVY21). The bbox (bounding box) indicates the spatial extent of the geospatial data.

### 8.3 Drawing Proportional Symbol Map
To create an interactive proportional symbol map in R, the view mode of tmap will be used.

The code below will turn on the interactive mode of tmap.

Interactive plot:
```{r}
tmap_mode("view")
```
#### 8.3.1 Starting with an interactive point symbol map
The code below is used to create an interactive point symbol map.

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "blue",
           size = 1,
           border.col = "green",
           border.lwd = 1)
```
#### 8.3.2 Making it proportional

To create a proportional symbol map, a numerical variable must be assigned to the size visual attribute. In the code, *Gp1Gp2Winnings* is used to define symbol size based on winnings.

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "blue",
           size = "Gp1Gp2 Winnings",
           border.col = "green",
           border.lwd = 1)
```
#### 8.3.3 Using a different color
The proportional symbol map can be enhanced by adding color as a visual attribute. In the code, `OUTLET_TYPE` is assigned to the color attribute, distinguishing different outlet types.

```{r}
tm_shape(sgpools_sf)+
tm_bubbles(col = "OUTLET TYPE", 
          size = "Gp1Gp2 Winnings",
          border.col = "black",
          border.lwd = 1)
```
#### 8.3.4 Facetd plots using `tm_facets()` method
The sync argument in `tm_facets()` allows multiple maps to be displayed with synchronized zoom and pan settings, ensuring a consistent view across all facets.

```{r}
tm_shape(sgpools_sf) +
  tm_bubbles(col = "OUTLET TYPE", 
          size = "Gp1Gp2 Winnings",
          border.col = "black",
          border.lwd = 1) +
  tm_facets(by= "OUTLET TYPE",
            nrow = 1,
            sync = TRUE)
```
We will switch **tmap’s** Viewer back to plot mode by using the code below.

Static plot:
```{r}
tmap_mode("plot")
```
### 8.4 References

-   Kam, T.S(2024). [Visual Statistical Analysis.](https://r4va.netlify.app/chap22)

### 8.5 Takeaway

::: callout-tip
## Key takeaways
-   Proportional symbol maps effectively visualize magnitude differences.
-   Color and facets enhance interpretation of categorical and quantitative data.
-   Interactive maps in tmap improve user exploration and comparison.
:::
